Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-01 index.qmd
section:background,dag; process:write

title: "Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism"

@@

<!-- 2. in what has been applied on?, and where is the gap? -->
DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by @Pollitt_2012b, @Jones_2015, @vanDaal_et_al_2016, @Bartholomew_et_al_2018, @Lesterhuis_2018, @Bartholomew_et_al_2020, and @Marshall_et_al_2020. However, its application transcends education, as exemplified by @Boonen_et_al_2020. The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992]. Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020]. Nevertheless, despite the method's widespread use, existing literature lacks a clear representation of the plausible mechanisms through which DCJ data could be generated. Particularly, there is no depiction of the complexity and the assumptions underpinning the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.

@@

<!-- 4. Solution to the gap -->
Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome. This study aims to fill this gap by representing DCJ within the framework of causal inference. Specifically, utilizing the structural approach [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system. The study then translates the scientific model into a probabilistic statistical model. This model aims to produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.

<!-- 5. final objective and relevance -->
Ultimately, this research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927]. Consequently, its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.

@@

This section introduces fundamental concepts in causal inference, such as directed acyclic graphs (DAGs), structural causal models (SCMs), and the flow of association and causation in graphs. It is not a comprehensive description of causal inference methods. Readers interested in deeper exploration should consult introductory papers like @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020. They may also find introductory books such as @Pearl_et_al_2018, @Neal_2020 and @McElreath_2020 useful. For more advanced study, seminal intermediate papers like @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books such as @Pearl_2009, @Morgan_et_al_2014 and @Hernan_et_al_2020 are recommended.

@@

<!-- 2. what are counterfactuals? -->
Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. According to @Pearl_et_al_2018, counterfactuals form the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association. Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018]. They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances."

@@

<!-- 1. Graph theory and the use of graphs -->
Graph theory is a branch of mathematics focused on the study of graphs. Graphs are mathematical structures that model pairwise relations between objects. They can represent physical relations, such as electrical circuits and roadways, and less tangible structures, such as ecosystems and sociological relations. Graphs have proven useful in various fields, including computer science, operations research, and the natural and social sciences [@Gross_et_al_2018].

@@

<!-- 3. How do we draw DAGs? and what are the building blocks of DAGs? -->
One key advantage of DAGs is that, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. @fig-dags illustrates these building blocks. @fig-dag_bb1 depicts two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related. @fig-dag_bb2 shows two connected nodes, illustrating a scenario where a parent node $X_{1}$ exerts a causal influence on a child node $X_{2}$. Consequently, $X_{2}$ is considered a *descendant* of $X_{1}$. @fig-dag_bb3 depicts a *chain* (or *pipe*), where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$. In this configuration, $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ a parent node of $X_{3}$. Furthermore, the DAG shows that $X_{1}$ is an *ancestor* of $X_{3}$ and that the relationship between these variables is entirely mediated by $X_{2}$. @fig-dag_bb4 illustrates a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$. In this scenario $X_{2}$ is a parent node of both $X_{1}$ and $X_{3}$. Finally, @fig-dag_bb5 depicts a *collider*, also known as *inmorality*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$. In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$.

@@

<!-- 4. A motivating example -->
Given the heuristic nature of DAGs, a motivating example can illustrate how these five fundamental building blocks can be used to construct the causal structure of a system. This example can also help clarify additional rules and conventions for drawing DAGs.

Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs to be investigated. Additionally, the problem indicates that a variable $X$ potentially influences both $Y$ and $T$. Such scenarios are not hard to imagine. For instance, $T$ could describe different treatments potentially influencing the binary onset of cancer ($Y$) while $X$ could denote the age of a patient. Using an example from DCJ, [we need an example]{style="color:red;"}. @fig-example presents two graphs describing the plausible causal structure of this system, the *simplified* and the *magnified* DAG.

<!-- 5. DAG for the motivating example -->
<!-- @fig-dag_example1 shows the endogenous variables $V=\{X,Z,Y\}$ as circled black nodes, indicating these variables are observed. The arrows connecting the variables indicate the direction of causal influence, while a lack of influence is indicated by the absence of arrows. Moreover, the exogenous variables $E=\{e_{X},e_{Z},e_{Y}\}$ represent everything else that is chosen not to be modeled explicitly. These exogenous variables, or *disturbances*, are usually represented by open circles to indicate their unobserved nature. Although this DAG, also known as Magnified DAG, explicitly shows the exogenous variables, conventionally these are omitted for brevity, resulting in an equivalent graph as shown in @fig-dag_example2 [@Cinelli_et_al_2020]. Sometimes it is convenient to show magnified graphs to show porblems in conditioning. -->

<!-- emphasize that everything can be observed -->
<!-- point out that the disturbances are usually stochastic in nature -->

@@

@@

<!-- commands for d-separation -->
<!-- \newcommand{\dsep}{\perp\!\!\!\perp} -->
<!-- \newcommand{\ndsep}{\not\!\perp\!\!\!\perp} -->

>>>>

title: "Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism"

@@

<!-- 2. in what has been applied on?, and where is the gap? -->
DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by @Pollitt_2012b, @Jones_2015, @vanDaal_et_al_2016, @Bartholomew_et_al_2018, @Lesterhuis_2018, @Bartholomew_et_al_2020, and @Marshall_et_al_2020. However, its application transcends education, as exemplified by @Boonen_et_al_2020. The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992]. Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020; @Bouwer_et_al_2023]. Nevertheless, despite the method's widespread use, existing literature lacks a clear representation of the plausible mechanisms through which DCJ data could be generated. Particularly, there is no depiction of the complexity and the assumptions underpinning the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.

@@

<!-- 4. Solution to the gap -->
Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021; @Bouwer_et_al_2023], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome. This study aims to fill this gap by representing DCJ within the framework of causal inference. Specifically, utilizing the structural approach [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system. The study then translates the scientific model into a probabilistic statistical model. This model aims to produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.

<!-- 5. final objective and relevance -->
Ultimately, this study provides a robust causal and probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927]. Consequently, its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.

@@

This section introduces fundamental concepts in causal inference, such as directed acyclic graphs (DAGs), structural causal models (SCMs), and the flow of association and causation in graphs. The section is not a comprehensive description of causal inference methods. Readers interested in deeper exploration should consult introductory papers like @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020. They may also find introductory books such as @Pearl_et_al_2018, @Neal_2020 and @McElreath_2020 useful. For more advanced study, seminal intermediate papers like @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books such as @Pearl_2009, @Morgan_et_al_2014 and @Hernan_et_al_2020 are recommended.

@@

<!-- 2. what are counterfactuals? -->
Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. According to @Pearl_et_al_2018, counterfactuals are the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association. Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018]. They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances."

@@

<!-- 1. Graph theory and the use of graphs -->
Graph theory is a branch of mathematics focused on the study of graphs. Graphs are mathematical structures modeling pairwise relations between objects. They can represent physical relations, such as electrical circuits and roadways, and less tangible structures, such as ecosystems and sociological relations. Graphs have proven useful in various fields, including computer science, operations research, and the natural and social sciences [@Gross_et_al_2018].

@@

<!-- 3. How do we draw DAGs? and what are the building blocks of DAGs? -->
One key advantage of DAGs is that, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. @fig-dags illustrates these blocks. @fig-dag_bb1 depicts two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related. @fig-dag_bb2 shows two connected nodes, illustrating a scenario where a parent node $X_{1}$ exerts a causal influence on a child node $X_{2}$. Consequently, $X_{2}$ is considered a *descendant* of $X_{1}$. @fig-dag_bb3 depicts a *chain* (or *pipe*), where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$. In this configuration, $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ a parent node of $X_{3}$. Furthermore, the DAG shows that $X_{1}$ is an *ancestor* of $X_{3}$ and that the relationship between these variables is entirely *mediated* by $X_{2}$. @fig-dag_bb4 illustrates a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$. In this scenario $X_{2}$ is a parent node of both $X_{1}$ and $X_{3}$. Finally, @fig-dag_bb5 depicts a *collider*, also known as *inmorality*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$. In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$.

@@

<!-- 4. A motivating example -->
Given the heuristic nature of DAGs, a motivating example can help illustrate the use of the five fundamental building blocks to construct a system's causal structure. This example can also help to clarify additional conventions for drawing DAGs.

Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs to be investigated. Additionally, the problem suggests that a variable $X$ potentially influences both $T$ and $Y$. Such scenarios are not hard to imagine. For instance, $T$ might represent different treatments that could affect the recovery from cancer $Y$, while $X$ could denote the patient's age. Similarly, in the context of a DCJ study like the one described by @Boonen_et_al_2020, $T$ could represent the duration of a child's cochlear implant use, which might influence the child's overall speech quality $Y$, with $X$ indicating the child's hearing status. Note that these descriptions do not specify the type of data or the functional forms of the variables, highlighting the heuristic nature of DAGs. [(not a bad example, but I prefer one using writing skills)]{style="color:blue;"}

<!-- 5. DAG for the motivating example -->
@fig-example presents two graphs illustrating the plausible causal structure of the motivating example. @fig-dag_example1 shows the *simplified* DAG. In this graph, the variables of interest or *endogenous* variables $V=\{T,X,Y\}$ are depicted as solid black circles, indicating that they are observed variables. The arrows in the graph reflect the expected direction of causal influences among the variables. @fig-dag_example2, in contrast, depicts the *magnified* version of the DAG. This graph includes, in addition to the endogenous variables, the *exogenous* variables $E=\{e_{T},e_{X},e_{Y}\}$. These exogenous variables are commonly referred to as *disturbances* or *errors*, because they represent factors not explicitly modeled. They are depicted as open circles, indicating their unobserved nature. While such exogenous variables are often omitted for simplicity, including them in the graph can be advantageous in certain scenarios, as they can help to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts that will be discussed in the next section.

<!-- 6. But where are the building blocks? -->
<!-- commands for d-separation -->
\newcommand{\dsep}{\perp\!\!\!\perp}
\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}

Furthermore, a detailed examination of both figures reveals the presence of at least four of the five fundamental building blocks. @fig-dag_example1 displays multiple descendants, evident in pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. It also illustrates a fork with $X \rightarrow \{T,Y\}$. Similarly, @fig-dag_example2 features multiple two-unconnected nodes, evident in the pairwise relations $\{e_{T} \dsep e_{X}\}$, $\{e_{T} \dsep e_{Y}\}$, and $\{e_{X} \dsep e_{Y}\}$, as well as colliders such as $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$. The symbol $\dsep$ denotes a concept know as *d-separation*, which roughly implies the independence of the variables. This concept will be introduced in the next section.

@@

<!-- point out that the disturbances are usually stochastic in nature -->

@@
